Outline paper - Multiple imputation for statistical disclosure control: Creating synthetic datasets with *mice*

### 2021-09-26 mice_synthesizing/paper.Rmd
Proceed with simulation section (not entirely finished yet).

An alternative solution has been proposed separately by @rubin_statistical_disclosure_1993 and @little_statistical_1993. Although their approaches differ to some extent, the overarching procedure is to use bonafide observed data to generate multiply imputed synthetic data sets that can be freely disclosed. While in practice, one could see this as replacing the observed data values by multiple draws from the posterior predictive distribution of the observed data, based on some imputation model, Rubin would argue that these synthetic data values are merely draws from the same true data generating model. In that sense, the observed data is never replaced, but the population is resampled from the information captured in the (incomplete) sample. Using this approach, the researcher could replace the observed data set as a whole with multiple synthetic versions. Alternatively, the researcher could opt to only replace a subset of the observed data. For example, one can choose to only replace dimensions in the data that could be compared with publicly available data sets or registers. Likewise, synthetisation could be limited to those values that are disclosive, such as high incomes or high turnovers. <!-- I don't think so, because the synthetic values can only be linked to observed values from other sources if they are exactly equal, or if they are the only value within a certain range. The latter might be possible when the number of values is very small, or when the observed value is so extreme that synthetic versions will still be so extreme. However, if the other values are synthetic as well, it is quite likely that synthetic information does not apply to the real-world observation. I think that the highest risk is in the possibility to extract a range of possible values, which might be uniquely match some other record that is available. -->

Conceptually, the synthetic data framework is based upon the building blocks of multiple imputation of missing data, as proposed by @rubin_multiple_1987. Instead of replacing just the missing values with multiple draws from the posterior predictive distribution, one could easily *overimpute* any observed sensitive values. Similarly to multiple imputation of missing data, the multiple synthetic datasets allow for correct statistical inferences, **regardless of the fact that the analyses do not use the "true" value**. The analyses over multiple synthetic data sets should be pooled into a single inference, so that the researcher can draw valid conclusions from the pooled results. To that respect, the variance should reflect the added variability that is induced by the imputation procedure.

Potentially, this approach could fulfill the needs for openly accessibly data, without running into barriers with regard to privacy and confidentiality constraints. However, there is no such thing as a free lunch: data collectors have to put effort in creating high-quality synthetic data. Also, the quality of the synthetic data is highly dependent on the imputation models, and using flawed models to generate synthetic data might bias subsequent analyses. Conversely, if the models used to create the synthetic data are able to preserve the relationships between the variables as in the original data, the synthetic data can be nearly as informative as the observed data. Thus, to fully exploit the benefits of synthetic data, the effort to actually create these high-quality data sets should be kept at a minimum.

To mitigate additional effort of creating synthetic data sets on behalf of the researcher, software aimed at multiple imputation of missing data can be employed. Especially if researchers used this software at an earlier stage in the research process, or acquired familiarity with it during earlier projects, the additional burden of creating synthetic data sets is relatively small. The R-package `mice` [@mice] implements multiple imputation of missing data in a straightforward and user-friendly manner. However, the functionality of `mice` is not restricted to the imputation of missing data, but allows to impute any value in the data: even observed values. Consequently, `mice` can be utilized for the creation of multiply imputed synthetic data sets.

After creating the multiply imputed synthetic data sets, the goal is to obtain valid statistical inferences in the spirit of @rubin_multiple_1987 and @neyman1934. In the missing data framework, this is done by performing statistical analyses on all imputed datasets, and pooling the results of the analyses according to Rubin's rules [@rubin_multiple_1987, pp.76]. In the synthetic data framework, the same procedure is followed, but with a slight twist: there are no values that remain constant over the synthetic data sets. The procedure of drawing valid inferences from multiple synthetic data sets is therefore slightly different.

@@

In `mice`, the generation of multiply imputed data sets to solve for unobserved values is straightforward. The following pseudocode details the multiple imputation of the `mice::boys` data set [@fredriks_boys_2000] into the object `imp` with `m = 10` imputated sets and `maxit = 7` iterations for the algorithm to converge, using the default imputations methods for each column data class.

@@

For example, overimputing synthetic values for both the observed and missing cells in the `mice::boys` data set into the object `syn`, given the same imputation parameters as before, can be realized by the following code execution.

@@

Choosing an adequate imputation model to impute the data is paramount, as a flawed imputation model may drastically impact the validity of inferences. Imputation models should be as flexible as possible to capture most of the patterns in the data, and to model possibly unanticipated data characteristics [@murray_multiple_2018; @rubin_18years_1996]. Parametric methods, albeit easy to implement in practice, may be too restrictive to capture generally complex patterns in the data, especially in the case of nonlinear relations and interactions between multiple variables. Classification and regression trees [CART; @breiman_cart_1984] allow to model more complex patterns in the data, and have therefore been suggested as an appropriate imputation method [@reiter_cart_2005; @burgette_reiter_cart_2010; @doove_buuren_recursive_2014]. Loosely speaking, CART sequentially splits the predictor space into non-overlapping regions in such a way that the within-region variance is as small as possible after every split. As such, CART does not impose any parametric distribution on the data, making it a widely applicable method that allows for a large variety of relationships within the data [@islr_2013]. Given these appealing characteristics and the call for the use of flexible methods when multiply imputing data, we will focus our illustrations and evaluations of `mice` to method `mice.impute.cart()`, realized by

@@

@@

[[[MOVE UPCOMING TWO PARAGRAPHS TO SIMULATION SECTION]]]

Additional parameters that come with the use of `mice.impute.cart()` are the complexity parameter `cp` and the minimum number of observations in any terminal node `minbucket`, that both restrain the flexibility of the imputation model. Appropriate values for these parameters, as well as the input for the `predictorMatrix`, depend on the data at hand. Hence, the specification of these parameters will be illustrated using an actual dataset. The `boys` data [@fredriks_boys_2000] that is in-built within the `mice` package is used to illustrate the specification of these parameters accordingly. **DETERMINE WHETHER THIS DATASET SUFFICES; BECAUSE IMPUTATION MIGHT MAKE THE DATA ARTIFICIALLY SIMILAR BEFORE SYNTHESIZING, MAKING SYNTHESIZING TOO EASY** Unfortunately, this dataset does not differ from the vast majority of collected datasets, in the sense that it suffers from missing data. For simplicity, the data is completed using the default `mice` imputation model to create a single, fully observed dataset. The data consists of the values of $`r nrow(mice::boys)`$ Dutch boys on the following $`r ncol(mice::boys)`$ variables: age (in years), height (in cm), weight (in kg), body mass index (BMI), head circumference (in cm), Genital Tanner stage (G1-G5), testicular volume (in ml) and region.

The values of the parameters `cp` and `minbucket` ought to adhere to the call for imputation models that are as flexible as possible. Accordingly, the complexity parameter is fixed at `cp = 1e-16 [[[CHECK THIS]]]` rather than the default value `1e-04` and the minimum number of observations in each terminal node is set at `minbucket` $= 3$ rather than the default value $5$. By allowing for more complexity in the imputation model, bias in the estimates from the synthetic dataset is reduced. The predictor matrix was specified so that BMI was not imputed from a model, but calculated from the imputed height and weight values ($BMI = kg / m^2$), an approach called passive imputation. Additionally, BMI is excluded as a predictor for weight and height, because BMI values are directly dependent on the values of weight and height. All other variables $Y_{-j}$ are used as predictors for variable $Y_j$. Additionally, the number of imputations is set to $m = 5$.

@@

To display the applicability of `mice` for generating synthetic data, we will display three distinct simulation set-ups that build up in complexity. The first case considers a multivariable regression model, in which the normally distributed outcome variable $y$ is regressed on three normally distributed predictor variables $\{x_1, x_2, x_3\}$, which are bivariately correlated with $\rho_{x_i, x_j} = 0.3$. The regression coefficients are specified in such a way that $\beta_2$ equals $2\beta_1$ and $\beta_3$ equals $3\beta_1$. Additionally, the proportion explained variance in $y$ equals $R^2=0.25$. Hence, the population regression weights equal $\beta_1 = 0.11$, $\beta_2 = 0.22$ and $\beta_3 = 0.33$.

The second simulation involves a logistic regression model, in which $y$ follows a Bernoulli distribution with probability 
$$
\frac{1}{1 + \text{exp}\{-X\beta\}},
$$
with predictor matrix $X$, consisting of 4 predictors. Here, $x_3$ moderates the relationship between $x_2$ and $y$, while $x_4$ is transformed into a categorical variable with three levels.

@@

The `mice` package has been developed originally for multiple imputation of missing data, but can be easily extended to the synthetic data framework. It proceeds by drawing missing or synthetic values from the joint distribution of the observed data. This joint density is generally hard or impossible to derive, which can be overcome by employing fully conditional specification (FCS). Consider a $n \times k$ data matrix $Y$, with values to be replaced $Y^{old}$, the values to be kept $Y^{keep}$ and the values that are already replaced $Y^{new}$, under the assumption that there are no missing values in the observed data set. Instead of drawing from the joint distribution, we draw from $k$ univariate conditional densities for each variable. That is, we draw from the distribution $P\big{(}Y_j^{old} | Y^{keep}, Y^{new}, Y^{old}_{-j}, \theta \big{)}$, where the drawn values are conditional upon the values that are to be kept, the synthetic values that are drawn at an earlier iteration and the values that still are to be replaced, with $j = 1, 2, \dots, k$ and $\theta$ indicating the parameter specifying the distribution of $Y_j^{old}$. Thus, the synthetic values are generated variable by variable. This procedure is executed $i = 1, 2, \dots, m$ times, resulting in $\textbf{D} = \{D^{(1)}, \dots, D^{(m)}\}$ synthetic datasets, with $D^{(i)} = (Y^{keep}, Y^{i, new})$.

@@

 - Explain algorithm (FCS) in more depth
 - What is synthetic data
 - Why is it useful
 - Why should it be easy to generate
 - Parrallel to missing data
 - If similar to missing data then why not use mice
 - All fun and game, but statistical properties should be intact
 - In this paper we demonstrate a straigthforward workflow for generating synthetic data sets and we investigate how to obtain valid inferences from synthesized data.

>>>>

An alternative solution has been proposed separately by @rubin_statistical_disclosure_1993 and @little_statistical_1993. Although their approaches differ to some extent, the overarching procedure is to use bonafide observed data to generate multiply imputed synthetic datasets that can be freely disclosed. While in practice, one could see this as replacing the observed data values by multiple draws from the posterior predictive distribution of the observed data, based on some imputation model, Rubin would argue that these synthetic data values are merely draws from the same true data generating model. In that sense, the observed data is never replaced, but the population is resampled from the information captured in the (incomplete) sample. Using this approach, the researcher could replace the observed dataset as a whole with multiple synthetic versions. Alternatively, the researcher could opt to only replace a subset of the observed data. For example, one can choose to only replace dimensions in the data that could be compared with publicly available datasets or registers. Likewise, synthetisation could be limited to those values that are disclosive, such as high incomes or high turnovers. <!-- I don't think so, because the synthetic values can only be linked to observed values from other sources if they are exactly equal, or if they are the only value within a certain range. The latter might be possible when the number of values is very small, or when the observed value is so extreme that synthetic versions will still be so extreme. However, if the other values are synthetic as well, it is quite likely that synthetic information does not apply to the real-world observation. I think that the highest risk is in the possibility to extract a range of possible values, which might be uniquely match some other record that is available. -->

Conceptually, the synthetic data framework is based upon the building blocks of multiple imputation of missing data, as proposed by @rubin_multiple_1987. Instead of replacing just the missing values with multiple draws from the posterior predictive distribution, one could easily *overimpute* any observed sensitive values. Similarly to multiple imputation of missing data, the multiple synthetic datasets allow for correct statistical inferences, **regardless of the fact that the analyses do not use the "true" value**. The analyses over multiple synthetic datasets should be pooled into a single inference, so that the researcher can draw valid conclusions from the pooled results. To that respect, the variance should reflect the added variability that is induced by the imputation procedure.

Potentially, this approach could fulfill the needs for openly accessibly data, without running into barriers with regard to privacy and confidentiality constraints. However, there is no such thing as a free lunch: data collectors have to put effort in creating high-quality synthetic data. Also, the quality of the synthetic data is highly dependent on the imputation models, and using flawed models to generate synthetic data might bias subsequent analyses. Conversely, if the models used to create the synthetic data are able to preserve the relationships between the variables as in the original data, the synthetic data can be nearly as informative as the observed data. Thus, to fully exploit the benefits of synthetic data, the effort to actually create these high-quality datasets should be kept at a minimum.

To mitigate additional effort of creating synthetic datasets on behalf of the researcher, software aimed at multiple imputation of missing data can be employed. Especially if researchers used this software at an earlier stage in the research process, or acquired familiarity with it during earlier projects, the additional burden of creating synthetic datasets is relatively small. The R-package `mice` [@mice] implements multiple imputation of missing data in a straightforward and user-friendly manner. However, the functionality of `mice` is not restricted to the imputation of missing data, but allows to impute any value in the data: even observed values. Consequently, `mice` can be utilized for the creation of multiply imputed synthetic datasets.

After creating the multiply imputed synthetic datasets, the goal is to obtain valid statistical inferences in the spirit of @rubin_multiple_1987 and @neyman1934. In the missing data framework, this is done by performing statistical analyses on all imputed datasets, and pooling the results of the analyses according to Rubin's rules [@rubin_multiple_1987, pp.76]. In the synthetic data framework, the same procedure is followed, but with a slight twist: there are no values that remain constant over the synthetic datasets. The procedure of drawing valid inferences from multiple synthetic datasets is therefore slightly different.

@@

In `mice`, the generation of multiply imputed datasets to solve for unobserved values is straightforward. The following pseudocode details the multiple imputation of the `mice::boys` dataset [@fredriks_boys_2000] into the object `imp` with `m = 10` imputated sets and `maxit = 7` iterations for the algorithm to converge, using the default imputations methods for each column data class.

@@

For example, overimputing synthetic values for both the observed and missing cells in the `mice::boys` dataset into the object `syn`, given the same imputation parameters as before, can be realized by the following code execution.

@@

Choosing an adequate imputation model to impute the data is paramount, as a flawed imputation model may drastically impact the validity of inferences. Imputation models should be as flexible as possible to capture most of the patterns in the data, and to model possibly unanticipated data characteristics [@murray_multiple_2018; @rubin_18years_1996]. Parametric methods, albeit easy to implement in practice, may be too restrictive to capture generally complex patterns in the data, especially in the case of nonlinear relations and interactions between multiple variables. Classification and regression trees [CART; @breiman_cart_1984] allow to model more complex patterns in the data, and have therefore been suggested as an appropriate imputation method [@reiter_cart_2005; @burgette_reiter_cart_2010; @doove_buuren_recursive_2014]. Loosely speaking, CART sequentially splits the predictor space into non-overlapping regions in such a way that the within-region variance is as small as possible after every split. As such, CART does not impose any parametric distribution on the data, making it a widely applicable method that allows for a large variety of relationships within the data [@islr_2013]. Given these appealing characteristics and the call for the use of flexible methods when multiply imputing data, we will focus our illustrations and evaluations of `mice` to method `mice.impute.cart()`, realized by:

@@

In a nutshell, the above code shows the simplicity of creating $m = 10$ synthetic datasets using `mice`. In practice, however, one should take some additional complicating factors into account. For example, one should account for deterministic relations in the data. Additionally, relations between variables may be described best using a different model than `CART`. Such factors are data dependent and should be considered by the imputer. In the next section, we will describe how the `boys` data can be adequately imputed. Additionally, we will show through simulations that this approach yields valid inferences.

We will show the suitability of using `mice` for synthesization using a simulation study, using the `boys` data. This dataset consists of the values of $`r nrow(mice::boys)`$ Dutch boys on the following $`r ncol(mice::boys)`$ variables: age (in years), height (in cm), weight (in kg), body mass index (BMI), head circumference (in cm), Genital Tanner stage (G1-G5), testicular volume (in ml) and region. Unfortunately, this dataset does not differ from the vast majority of collected datasets, in the sense that it suffers from missing data. For simplicity, the data is completed using the default `mice` imputation model for all predictors except BMI, which is imputed using a passive imputation model due to the deterministic relation with weight and height ($\text{BMI} = \frac{\text{weight}}{\text{height (in meters)}^2}$) to create a single, fully observed dataset:

```{r, include = FALSE}
library(mice)
library(magrittr)
library(tidyverse)
library(furrr)
library(knitr)
library(kableExtra)
```

```{r, results = FALSE, message = FALSE, warning = FALSE}
# create a single imputed, completely observed `boys` dataset
set.seed(123)

meth <- make.method(boys)
meth["bmi"] <- "~ I(wgt / (hgt / 100)^2)"
pred <- make.predictorMatrix(boys)
pred[c("hgt", "wgt"), "bmi"] <- 0

imp <- mice(boys, 
            m = 1,
            maxit = 10,
            method = meth,
            predictorMatrix = pred)

data <- complete(imp)
```

To induce an appropriate amount of sampling variance, 1000 bootstrap samples of the `boys` data will be synthesized with $m = 5$ imputations. The data will be synthesized using the `CART` imputation method. Additional parameters that come with the use of `mice.impute.cart()` are the complexity parameter `cp` and the minimum number of observations in any terminal node `minbucket`, that both constrain the flexibility of the imputation model. The values of the parameters `cp` and `minbucket` ought to adhere to the call for imputation models that are as flexible as possible. Appropriate values for these parameters, as well as the input for the `predictorMatrix`, depend on the data at hand. In the current example, the complexity parameter is specified at `cp = 1e-08` rather than the default value `1e-04`, and the minimum number of observations in each terminal node is set at `minbucket` $= 3$ rather than the default value $5$. By allowing for more complexity in the imputation model, bias in the estimates from the synthetic dataset is reduced. Additionally, since the missingness pattern is monotone, the number of iterations can be set to `maxit = 1`.

```{r eval = FALSE}
syn <- mice(boys,
            m = 10,
            maxit = 7,
            method = "cart",
            where = matrix(TRUE,
                           nrow = nrow(boys),
                           ncol = ncol(boys)),
            cp = 1e-08,
            minbucket = 3)
```

@@

@@

**!!PROBABLY NOT USED!!**

*To display the applicability of `mice` for generating synthetic data, we will display three distinct simulation set-ups that build up in complexity. The first case considers a multivariable regression model, in which the normally distributed outcome variable $y$ is regressed on three normally distributed predictor variables $\{x_1, x_2, x_3\}$, which are bivariately correlated with $\rho_{x_i, x_j} = 0.3$. The regression coefficients are specified in such a way that $\beta_2$ equals $2\beta_1$ and $\beta_3$ equals $3\beta_1$. Additionally, the proportion explained variance in $y$ equals $R^2=0.25$. Hence, the population regression weights equal $\beta_1 = 0.11$, $\beta_2 = 0.22$ and $\beta_3 = 0.33$.*

*The second simulation involves a logistic regression model, in which $y$ follows a Bernoulli distribution with probability*
$$
\frac{1}{1 + \text{exp}\{-X\beta\}},
$$
*with predictor matrix $X$, consisting of 4 predictors. Here, $x_3$ moderates the relationship between $x_2$ and $y$, while $x_4$ is transformed into a categorical variable with three levels.*

@@

The `mice` package has been developed originally for multiple imputation of missing data, but can be easily extended to the synthetic data framework. It proceeds by drawing missing or synthetic values from the joint distribution of the observed data. This joint density is generally hard or impossible to derive, which can be overcome by employing fully conditional specification (FCS). Consider a $n \times k$ data matrix $Y$, with values to be replaced $Y^{old}$, the values to be kept $Y^{keep}$ and the values that are already replaced $Y^{new}$, under the assumption that there are no missing values in the observed dataset. Instead of drawing from the joint distribution, we draw from $k$ univariate conditional densities for each variable. That is, we draw from the distribution $P\big{(}Y_j^{old} | Y^{keep}, Y^{new}, Y^{old}_{-j}, \theta \big{)}$, where the drawn values are conditional upon the values that are to be kept, the synthetic values that are drawn at an earlier iteration and the values that still are to be replaced, with $j = 1, 2, \dots, k$ and $\theta$ indicating the parameter specifying the distribution of $Y_j^{old}$. Thus, the synthetic values are generated variable by variable. This procedure is executed $i = 1, 2, \dots, m$ times, resulting in $\textbf{D} = \{D^{(1)}, \dots, D^{(m)}\}$ synthetic datasets, with $D^{(i)} = (Y^{keep}, Y^{i, new})$.

@@

 - Explain algorithm (FCS) in more depth
 - What is synthetic data
 - Why is it useful
 - Why should it be easy to generate
 - Parrallel to missing data
 - If similar to missing data then why not use mice
 - All fun and game, but statistical properties should be intact
 - In this paper we demonstrate a straigthforward workflow for generating synthetic datasets and we investigate how to obtain valid inferences from synthesized data.
